---
title: "Gustavo"
author: "S. Schmitt"
date: last-modified
format: 
  html:
    embed-resources: true
    toc: true
bibliography: references.bib
---

# Lecture

-   About tree height allometry: @molto2014
-   TALLO database: @jucker2022
-   `cmdstanr` installation: <https://mc-stan.org/cmdstanr/articles/cmdstanr.html>
-   `stan` programming: <https://mc-stan.org/>

# Setup

You only need the `tidyverse` package for table manipulation and figures preparation, `cmdstanr`, `brms`, and `bayesplot` for Bayesian statistics.

```{r libs}
#| message: false
#| warning: false
library(tidyverse)
library(cmdstanr)
library(brms)
library(bayesplot)
```

# Data

Data are tree hight and diameter from the TALLO database [@jucker2022] itself using Dos Santos data for Jamari. Longitude and latitude are in degrees, so 1 degree is about 110 km at the equator, and we use a buffer of 10km.

```{r data}
#| message: false
#| warning: false
lon <- -62.99
lat <- -9.11
buffer <- 10/110
data <- read_csv("gustavo/tallo.csv") %>% 
  filter(longitude >= lon-buffer, longitude <= lon+buffer,
          latitude >= lat-buffer, latitude <= lat+buffer) %>%  
  separate(species, c("genus", "species")) %>% 
  select(genus, tree_id, height_m, stem_diameter_cm) %>% 
  na.omit() %>% 
  group_by(genus) %>% 
  filter(n() > 10) %>% 
  ungroup() %>% 
  mutate(genusnum = as.numeric(as.factor(genus)))
ggplot(data, aes(stem_diameter_cm, height_m)) +
  geom_point(alpha = 0.25) +
  theme_bw() +
  geom_smooth(method = "lm", formula = y ~ log(x), se = FALSE) +
  xlab("diameter (cm)") + ylab("height (m)") +
  facet_wrap(~ genus)
```

# Model

We model tree height $h$ in function of tree diameter $d$ as:

$$
\begin{align}
h_{g,u,i} \sim logN( log(\mu_g), \sigma) \\
\mu_g = \frac{\alpha_g\times d_{g,u,i}}{\beta_g + d_{G,u,i}} \\
\alpha_g \sim N(\alpha, \sigma_a) \\
\beta_g \sim N(\beta, \sigma_b) \\
u_u \sim N(0, \sigma_u)
\end{align}
$$

<!-- $$ -->
<!-- \begin{align} -->
<!-- h_{s,u,i} \sim logN( log(\mu_s) + u_u, \sigma) \\ -->
<!-- \mu_s = \frac{\alpha_s\times d_{s,u,i}}{\beta_s + d_{s,u,i}} \\ -->
<!-- u_u \sim N(0, \sigma_u) \\ -->
<!-- \begin{bmatrix} \alpha_s \\ \beta_s \end{bmatrix} \sim MVN(\begin{bmatrix} \alpha \\ \beta \end{bmatrix}],  -->
<!-- \begin{bmatrix} \sigma_a^2 & \rho \\ \rho & \sigma_s^2 \end{bmatrix}) \\ -->
<!-- \end{align} -->
<!-- $$ -->

with $h_{max}$ ...

# Example with `cmdstanr`

```{r stan_model}
mdata <- list(
  N = nrow(data),
  G = max(data$genusnum),
  h = data$height_m,
  d = data$stem_diameter_cm,
  genus = data$genusnum
)
model <- cmdstan_model("gustavo/gustavo.stan")
model
```

```{r stan_fit}
#| output: false
fit <- model$sample(data = mdata, parallel_chains = 4)
```

```{r}
fit
```

```{r }
fit$draws(c("alpha_g")) %>% 
  mcmc_intervals()
```

```{r}
fit$draws(c("beta_g")) %>% 
  mcmc_intervals()
```

```{r}
fit$draws(c("alpha", "beta", "sigma_a", "sigma_b", "sigma")) %>% 
  mcmc_hist()
```

```{r}
data %>%
  mutate(pred = fit$summary("mu")$median) %>% 
  ggplot(aes(pred, height_m)) +
  geom_abline() +
  geom_point() +
  theme_bw()
```

# References
