---
title: "Gustavo recovery"
author: "S. Schmitt"
date: last-modified
format: 
  html:
    embed-resources: true
    toc: true
bibliography: references.bib
---

# Lecture

-   About the model: @maurent2023
-   `cmdstanr` installation: <https://mc-stan.org/cmdstanr/articles/cmdstanr.html>
-   `stan` programming: <https://mc-stan.org/>

# Setup

You only need the `tidyverse` package for table manipulation and figures preparation, `cmdstanr` and `bayesplot` for Bayesian statistics.

```{r libs}
#| message: false
#| warning: false
library(tidyverse)
library(cmdstanr)
library(bayesplot)
```

# Model

We model $plot$ biomass $agb$ in function of time post disturbance $time$ as:

$$
\begin{align}
agb \sim \mathcal N [ agb_{0,plot} + (agb_{\infty,plot} - agb_{0,plot}) \times (1 - e^{-\lambda_{plot} \times time}), \sigma ] \\
agb_{\infty,plot} \sim \mathcal N (agb_{prelogging}, \sigma_{prelogging})\\
agb_{0,plot} = agb_{\infty,plot} - agb_{\infty,plot} \times \alpha_{plot} \\
\alpha_{plot} \sim \mathcal N(\alpha, \sigma_a) \\
\lambda_{plot} \sim \mathcal N(\lambda, \sigma_l)
\end{align}
$$

With:

-   $agb_{\infty}$ : biomass at equilibirum
-   $agb_{0}$ : biomass after disturbance (depending on $\alpha$ and $agb_{\infty}$)
-   $\alpha$ : disturbance intensity (% biomass)
-   $\lambda$ : recovery rate (time relative)

```{r fig}
#| message: false
#| warning: false
ltp <- function(time, alpha, agb_inf, lambda) {
  agb_0 <- agb_inf * alpha
  ltp <- 1 - exp(-lambda * time)
  agb_0 + (agb_inf - agb_0) * (ltp)
}
data.frame(time = seq(0.1, 50, length.out = 100)) %>%
  mutate(y_ltp = ltp(time, alpha = 0.5, agb_inf = 20, lambda = 0.05)) %>%
  bind_rows(data.frame(time = 0, y_stp = 20, y_ltp = 20)) %>%
  ggplot(aes(time)) +
  geom_line(aes(y = y_ltp)) +
  theme_bw() +
  annotate("text",
    x = 52, y = 19.5, label = "agb[infinity]",
    col = "#9ea1d4", parse = TRUE, size = 5
  ) +
  annotate("text",
    x = 20, y = 15, label = "lambda",
    col = "#a8d2d1", parse = TRUE, size = 5
  ) +
  annotate("text",
    x = -4, y = 15, label = "alpha",
    col = "#fc8a8a", parse = TRUE, size = 5
  ) +
  theme(axis.title = element_blank()) +
  geom_segment(aes(x = -2, y = 10, xend = -2, yend = 20),
    arrow = arrow(length = unit(0.2, "cm")), col = "#fc8a8a"
  ) +
  geom_segment(aes(x = -2, y = 20, xend = -2, yend = 10),
    arrow = arrow(length = unit(0.2, "cm")), col = "#fc8a8a"
  ) +
  theme(axis.title = element_blank()) +
  geom_segment(aes(x = -5, y = 20, xend = 0, yend = 20))
```

# Data

We are going to build fake data.

```{r data}
#| message: false
#| warning: false
n_plot <- 20
agb_inf <- 300
alpha <- 0.25
prelog <- 3
postlog <- 11
sigma_plot <- 25
sigma_year <- 5
lambda <- 0.1
data <- tibble(plot = paste0("PP", 1:n_plot)) %>% 
  mutate(type = rep(c("undisturbed", "logged"), each = n_plot/2)) %>% 
  mutate(agb_inf = rnorm(n_plot, agb_inf, sigma_plot)) %>% 
  mutate(agb_0 = ifelse(type == "logged", agb_inf - alpha*agb_inf, agb_inf)) %>% 
  mutate(time = list(1:(prelog+postlog)-prelog)) %>% 
  unnest(time) %>% 
  group_by(plot) %>% 
  mutate(agb = ifelse(time > 0, 
                      agb_0 + (agb_inf - agb_0) * (1 - exp(-lambda * time)),
                      agb_inf)) %>% 
  ungroup() %>% 
  mutate(agb_obs = rnorm(n(), agb, sigma_year)) %>% 
  mutate(plotnum = as.numeric(as.factor(plot)),
         typenum = as.numeric(as.factor(type)))
ggplot(data, aes(time, agb, col = type, group = plot)) +
  geom_vline(xintercept = 0) +
  geom_line() +
  geom_point(aes(y = agb_obs)) +
  theme_bw()
```

# Stan

The model writing is a bit complicated:

```{r}
model <- cmdstan_model("gustavo/gustavo_recovery.stan")
model$print()
```

We are going to focus on disturbed plots. The most complicated is to prepare the data. For that we must split prelogging and postlogging information:

```{r}
data <- filter(data, type == "logged")
data_pre <- filter(data, time <= 0)
data_rec <- filter(data, time > 0)
mdata <- list(
  N_rec = nrow(data_rec),
  N_pre = nrow(data_pre),
  P = max(data$plotnum),
  agb = data_rec$agb_obs,
  time = data_rec$time,
  agb_pre =  data_pre$agb_obs,
  plot =  data_rec$plotnum,
  plot_pre = data_pre$plotnum
)
```

```{r}
#| output: false
fit <- model$sample(data = mdata, parallel_chains = 4)
```

```{r}
fit$draws("lp__") %>% mcmc_trace()
```

```{r}
fit$draws("alpha_p") %>% mcmc_intervals() + theme_bw()
```

```{r}
fit$draws("agb_inf_p") %>% mcmc_intervals() + theme_bw()
```

```{r}
fit$draws("lambda_p") %>% mcmc_intervals() + theme_bw()
```

```{r}
#| warning: false
fit$draws(c("alpha", "lambda")) %>% mcmc_dens() + theme_bw()
```

```{r}
data_rec %>% 
  mutate(pred = fit$summary("mu")$median) %>% 
  ggplot(aes(pred, agb_obs)) +
  geom_abline() +
  geom_point() +
  theme_bw()
```

```{r}
data_rec %>% 
  mutate(pred = fit$summary("mu")$median) %>% 
  ggplot(aes(time, pred, group = plot)) +
  geom_point(aes(y = agb_obs)) +
  geom_line() +
  theme_bw()
```

# References
