---
title: "Phenology"
author: "Sylvain Schmitt"
date: last-modified
format: 
  html:
    embed-resources: true
    toc: true
    toc_float: true
    cache: true
---

# Setup

You only need the `̀tidyverse` package for table manipulation and figures preparation, `cmdstanr`, & `bayesplot` for Bayesian statistics, and `cowplot` for figures.

```{r libs}
#| message: false
#| warning: false
library(tidyverse)
library(dplyr)
library(cmdstanr)
library(bayesplot)
library(cowplot)
```

# Data

Available data include:

- photosynthetic efficiency measurement $phi$ (`phi2_0`)
- photosynthetically active radiation $par$ (`Light.Intensity..PAR.`)
- vapour pressure deficit $vpd$ (`VPD`)
- species $sp$ (`Especie_final`)
- ... (*among others*)

```{r data}
#| message: false
#| warning: false
#| fig-cap: "Caption."
data <- readxl::read_xlsx("dados_PAM_summary.xlsx")
g_par <- data %>% 
  ggplot(aes(Light.Intensity..PAR., phi2_0)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x) +
  theme_bw() +
  scale_y_log10()
g_vpd <- data %>% 
  ggplot(aes(VPD, phi2_0)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw()
g_age <- data %>% 
  ggplot(aes(n_days, phi2_0, col = Especie_final)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  theme_bw() +
  ylim(0, 1) +
  scale_color_discrete(guide = "none")
plot_grid(g_par, g_vpd, g_age, align = "hv")
```

# Model data

We prepare data for modelling adding a new column `speciesnum` with species as numerical indexes to the data, and adding to new variables $N$ the number of observations and $S$ the number of species.

```{r model_data}
data <- data %>%
  mutate(speciesnum = as.numeric(as.factor(Especie_final))) 
mdata <- list(
  N = nrow(data),
  S = max(data$speciesnum),
  phi = data$phi2_0,
  age = data$n_days,
  vpd = data$VPD,
  par = data$Light.Intensity..PAR.,
  sp = data$speciesnum,
  N_p = max(data$n_days)
)
```

# Model $par$

We used an exponential decay to represent the $par$ effect:

$$phi \sim \mathcal N (a_p \times exp(-b_p\times par), \sigma) $$

$a_p$ is controlling the intercept efficiency of $phi$ when par is null, while $b_p$ is controlling the slope of the exponential decay:

```{r par_fig}
expand_grid(par = 1:2500, a = 0.6, b_p = exp(seq(-7, 0))) %>% 
  mutate(phi = a+exp(-b_p*par)) %>% 
  ggplot(aes(par, phi, col = b_p, group = b_p)) +
  geom_line() +
  theme_bw() +
  scale_color_viridis_c()
```

```{r par_fit}
#| eval: false
model <- cmdstan_model("models/par.stan")
list.files("chains/par/", full.names = TRUE) %>% unlink()
fit <- model$sample(mdata, parallel_chains = 4, output_dir = "chains/par/")
```

Constraining parameters in the right space, the model converged well:

```{r par_fit_res}
fit <- as_cmdstan_fit(list.files("chains/par/", full.names = TRUE))
fit$draws(c("lp__", "ap", "bp_log", "sigma")) %>% 
  mcmc_trace() + theme_bw()
```

$par$ alone is explaining already 90% of the observed variation (Bayesian R-squared):

```{r par_pvo}
rmsep <- fit$summary("rmsep")
brsq <- fit$summary("brsq")
data %>% 
  mutate(pred_par = fit$summary("mu_par", median)$median) %>% 
  ggplot(aes(pred_par, phi2_0)) +
  geom_abline() +
  geom_point() +
  theme_bw() +
  annotate("text", 0.2, 0.7, 
           label = paste0("RMSEP=", round(rmsep$median, 3))) +
  annotate("text", 0.2, 0.65, 
           label = paste0("R2=", round(brsq$median, 2)))
```

The projection seems good but high values of par might have underestimated phi:

```{r par_proj}
fit$summary("mu_par") %>% 
  mutate(par = mdata$par) %>% 
  mutate(phi_obs = mdata$phi) %>% 
  ggplot(aes(par, median)) +
  geom_line(aes(y = q5), linetype = "dashed") +
  geom_line(aes(y = q95), linetype = "dashed") +
  geom_line() +
  geom_point(aes(y = phi_obs)) +
  theme_bw()
```

# Model $vpd$

We used a simple linear effect to represent the $vpd$ effect as we did not have much expectation and preliminary results indicated weak to no effect.

$$phi \sim \mathcal N (a_v+b_v\times vpd, \sigma) $$

```{r vpd_fit}
#| eval: false
model <- cmdstan_model("models/vpd.stan")
list.files("chains/vpd/", full.names = TRUE) %>% unlink()
fit <- model$sample(mdata, parallel_chains = 4, output_dir = "chains/vpd/")
```

The model converged well:

```{r vpd_fit_res}
fit <- as_cmdstan_fit(list.files("chains/vpd/", full.names = TRUE))
fit$draws(c("lp__", "av", "bv", "sigma")) %>% 
  mcmc_trace() + theme_bw()
```

$vpd$ alone is explaining only 15% of the observed variation (Bayesian R-squared):

```{r vpd_pvo}
rmsep <- fit$summary("rmsep")
brsq <- fit$summary("brsq")
data %>% 
  mutate(pred_par = fit$summary("mu_vpd", median)$median) %>% 
  ggplot(aes(pred_par, phi2_0)) +
  geom_abline() +
  geom_point() +
  theme_bw() +
  annotate("text", 0.3, 0.7, 
           label = paste0("RMSEP=", round(rmsep$median, 3))) +
  annotate("text", 0.3, 0.65, 
           label = paste0("R2=", round(brsq$median, 2)))
```

The projection seems bad, in accordance with the fact that most variation is explained by par that might covary with vpd and create the negative relation:

```{r vpd_proj}
fit$summary("mu_vpd") %>% 
  mutate(vpd = mdata$vpd) %>% 
  mutate(phi_obs = mdata$phi) %>% 
  ggplot(aes(vpd, median)) +
  geom_line(aes(y = q5), linetype = "dashed") +
  geom_line(aes(y = q95), linetype = "dashed") +
  geom_line() +
  geom_point(aes(y = phi_obs)) +
  theme_bw()
```

# Model $age$

Modelling the photosynthetic efficiency as a function of its $age$ is often carried out using Weibull-type functions or exponential decline functions, as photosynthesis reaches its maximum (peak) shortly after leaf expansion and then declines during senescence as:

$$phi \sim \mathcal N [ a_s \times \frac{age}{b_s}^{k_s-1} \times exp(\frac{age}{b_s}^{k_s}), \sigma]$$

With:

- $a$ : Scale parameter for each species (related to the magnitude of efficiency)
- $b$ : Time scale parameter for each species (related to the timing of senescence)
- $k$ : Shape parameter for each species (determines the curvature/peaking)

From this parameters we can derive to parameters of interest which are the peak timing and intensity $age_{opt,s}$ and $phi_{max,s}$ as:

- $age_{opt,s} = b_s \times(\frac{k_s-1}{k_s})^{\frac 1k_s}$
- $phi_{max,s} = a_s \times \frac{age_{opt,s}}{b_s}^{k_s-1}*exp(-\frac{age_{opt,s}}{b_s}^{k_s})$

```{r age_fig}
weibull <- function(t, a, b, k) 
  a*(t/b)^(k-1)*exp(-(t/b)^k)
expand_grid(k = seq(1, 2, 0.1),
            b = c(100, 200, 300),
            a = c(10, 20, 30),
            time = 0:365) %>% 
  mutate(p = weibull(time, a = a, b = b, k = k)) %>% 
  ggplot(aes(time, p, col = k, group = k)) +
  geom_line() +
  theme_bw() +
  xlab("Time [day]") + ylab("Photosynthetic capacity") +
  facet_grid(a ~ b, labeller = label_both) +
  scale_color_viridis_c()
```

```{r age_fit}
#| eval: false
model <- cmdstan_model("models/age.stan")
list.files("chains/age/", full.names = TRUE) %>% unlink()
fit <- model$sample(mdata, parallel_chains = 4, output_dir = "chains/age/")
```

The model converged ok:

```{r age_fit_res}
fit <- as_cmdstan_fit(list.files("chains/age/", full.names = TRUE))
fit$draws(c("lp__", "a", "b", "k", "sigma")) %>% 
  mcmc_trace() + theme_bw()
```

$age$ alone is explaining only 35% of the observed variation (Bayesian R-squared):

```{r age_pvo}
rmsep <- fit$summary("rmsep")
brsq <- fit$summary("brsq")
data %>% 
  mutate(pred_par = fit$summary("mu_age", median)$median) %>% 
  ggplot(aes(pred_par, phi2_0)) +
  geom_abline() +
  geom_point() +
  theme_bw() +
  annotate("text", 0.3, 0.7, 
           label = paste0("RMSEP=", round(rmsep$median, 3))) +
  annotate("text", 0.3, 0.65, 
           label = paste0("R2=", round(brsq$median, 2)))
```

The projection seems ok:

```{r age_proj}
fit$summary("mu_age") %>% 
  mutate(age = mdata$age) %>% 
  mutate(phi_obs = mdata$phi) %>% 
  mutate(species = mdata$sp) %>% 
  ggplot(aes(age, median)) +
  geom_line(aes(y = q5), linetype = "dashed") +
  geom_line(aes(y = q95), linetype = "dashed") +
  geom_line() +
  geom_point(aes(y = phi_obs)) +
  theme_bw() +
  facet_wrap(~ species)
```

# Full model

Now we are assembling the three in a single model:

$$phi \sim \mathcal N [ a_p \times exp(-b_p\times par) + b_v\times vpd + a_s \times \frac{age}{b_s}^{k_s-1} \times exp(\frac{age}{b_s}^{k_s}), \sigma]$$

```{r full_fit}
#| eval: false
model <- cmdstan_model("models/full.stan")
list.files("chains/full/", full.names = TRUE) %>% unlink()
fit <- model$sample(mdata, parallel_chains = 4, output_dir = "chains/full/")
```

The model converged well:

```{r full_fit_res}
fit <- as_cmdstan_fit(list.files("chains/full/", full.names = TRUE))
fit$draws(c("lp__", "ap", "bp_log", "bv", "a", "b", "k", "sigma")) %>% 
  mcmc_trace() + theme_bw()
```

The full model is explaining only 94% of the observed variation (Bayesian R-squared):

```{r full_pvo}
rmsep <- fit$summary("rmsep")
brsq <- fit$summary("brsq")
data %>% 
  mutate(pred_par = fit$summary("mu", median)$median) %>% 
  ggplot(aes(pred_par, phi2_0)) +
  geom_abline() +
  geom_point() +
  theme_bw() +
  annotate("text", 0.3, 0.7, 
           label = paste0("RMSEP=", round(rmsep$median, 3))) +
  annotate("text", 0.3, 0.65, 
           label = paste0("R2=", round(brsq$median, 2)))
```

Species projections are good, we can then look at the response parameters tmax and pmaw corresponding to timing and value of maximal efficiency:

```{r full_proj}
t <- fit$summary("y_sp") %>% 
  separate(variable, c("x1", "x2", "sp", "age")) %>% 
  mutate(age = as.numeric(age)*10)
t %>% 
  na.omit() %>% 
  ggplot(aes(age, median)) +
  geom_line(aes(y = q5), linetype = "dashed") +
  geom_line(aes(y = q95), linetype = "dashed") +
  geom_line() +
  theme_bw() +
  facet_wrap(~ sp)
```

The maximum timing varies from  123 to 574 days:

```{r full_tmax}
fit$draws("tmax_s") %>% 
  mcmc_areas_ridges() +
  theme_bw() +
  scale_x_sqrt()
```

The maximum values varies from 0.086 to 0.200:

```{r full_pmax}
fit$draws("pmax_s") %>% 
  mcmc_areas_ridges() +
  theme_bw()
```

We can make a biplot of the two:

```{r full_pmax_tmax}
id <- data %>% 
  select(Especie_final, speciesnum) %>% 
  unique()
fit$summary(c("tmax_s", "pmax_s")) %>% 
  separate(variable, c("parameter", "s", "speciesnum"), convert = TRUE) %>% 
  select(-s) %>% 
  left_join(id) %>% 
  gather(metric, value, -parameter, -speciesnum, -Especie_final) %>% 
  mutate(parameter_metric = paste0(parameter, "_", metric)) %>% 
  select(-parameter, -metric) %>% 
  pivot_wider(names_from = parameter_metric, values_from = value) %>% 
  ggplot(aes(tmax_median, pmax_median, col = Especie_final)) +
  geom_point() +
  geom_segment(aes(x = tmax_median - tmax_sd, xend = tmax_median + tmax_sd), alpha = .5) +
  geom_segment(aes(y = pmax_median - pmax_sd, yend = pmax_median + pmax_sd),
    alpha = .5
  ) +
  theme_bw() +
  scale_x_sqrt() +
  xlab("Peak timing [Tmax, days]") +
  ylab("Peak intensity [Pmax]") +
  scale_color_discrete("") +
  theme(legend.text = element_text(face = "italic"))
```


