---
title: "Phenology"
author: "Sylvain Schmitt"
date: last-modified
format: 
  html:
    embed-resources: true
    toc: true
    toc_float: true
---

# Setup

You only need the tidyverse package for table manipulation and figures preparation, cmdstanr, and bayesplot for Bayesian statistics.

```{r libs}
#| message: false
#| warning: false
library(tidyverse)
library(dplyr)
library(cmdstanr)
library(bayesplot)
```

# Weibull

## Model

Modelling the photosynthetic efficiency or capacity of a leaf as a function of its age (ageing) is often carried out using Weibull-type functions or exponential decline functions, as photosynthesis reaches its maximum (peak) shortly after leaf expansion and then declines during senescence.

Photosynthetic capacity $P$ along time $t$ can be expressed as:

$$
P(t) \sim \mathcal N [ a \times \frac{t}{b}^{k-1} \times exp(\frac{t}{b}^{k}), \sigma]
$$
With:

* $a$ : Scale parameter (related to the magnitude of efficiency)
* $b$ : Time scale parameter (related to the timing of senescence)
* $k$ : Shape parameter (determines the curvature/peaking)

```{r}
a <- 30
b <- 300
k <- 1.5
weibull <- function(t, a, b, k) 
  a*(t/b)^(k-1)*exp(-(t/b)^k)
tmax <- b*((k-1)/k)^(1/k)
pmax <- weibull(tmax, a, b, k)
tibble(t = 0:365) %>% 
  mutate(p = weibull(t, a, b, k)) %>% 
  ggplot(aes(t, p)) +
  geom_line() +
  theme_bw() +
  xlab("Time [day]") + ylab("Photosynthetic capacity") +
  geom_vline(xintercept = tmax, linetype = "dashed") +
  geom_hline(yintercept = pmax, linetype = "dashed")
```

```{r}
expand_grid(k = seq(1, 2, 0.1),
            b = c(100, 200, 300),
            a = c(10, 20, 30),
            time = 0:365) %>% 
  mutate(p = weibull(time, a = a, b = b, k = k)) %>% 
  ggplot(aes(time, p, col = k, group = k)) +
  geom_line() +
  theme_bw() +
  xlab("Time [day]") + ylab("Photosynthetic capacity") +
  facet_grid(a ~ b, labeller = label_both)
```

## Data

```{r}
n <- 100
a <- 30
b <- 300
k <- 1.5
sigma <- 2
data <- tibble(age = sample(1:365, n)) %>% 
  mutate(p = rnorm(n(), weibull(age, a, b, k), sigma))
ggplot(data, aes(age, p)) +
  geom_point() +
  theme_bw()
```

## Fit

```{r}
model <- cmdstan_model("weibull.stan")
model$print()
```

```{r}
#| output: false
mdata <- list(
  N = nrow(data),
  t = data$age,
  p = data$p
)
fit <- model$sample(mdata, parallel_chains = 4)
```

```{r}
fit$draws(c("lp__", "a", "b", "k")) %>% mcmc_trace() + theme_bw()
```

```{r}
#| message: false
#| warnign: false
fit$draws(c("tmax", "pmax")) %>% mcmc_dens() + theme_bw()
```

```{r}
data %>% 
  mutate(pred = fit$summary("mu", median)$median) %>% 
  ggplot(aes(pred, p)) +
  geom_abline() +
  geom_point() +
  theme_bw()
```

```{r}
data %>% 
  mutate(pred = fit$summary("mu", median)$median) %>% 
  ggplot(aes(age, p)) +
  geom_line(aes(y = pred)) +
  geom_point() +
  theme_bw()
```

# Breakpoint

## Model

We could also use a breakpoint model (often referred to as the Broken Stick or Change Point Model) to estimate at what point (leaf age) the slope of photosynthetic efficiency changes radically.

Basically, before the breakpoint $\tau$:

$$
P(t) \sim \mathcal N [ a_1 + b_1 \times t, \sigma]
$$

and after the breakpoint $\tau$:

$$
P(t) \sim \mathcal N [ a_2 + b_2 \times t, \sigma]
$$

With:

* $b_1$ : the slope before breakpoint
* $b$ : Time scale parameter (related to the timing of senescence)
* $b_2$ : the slope after breakpoint

```{r}
a <- 0
b1 <- 0.1
tau <- 200
b2 <- -0.05
breakpoint <- function(t, tau, a, b1, b2){
  if(t < tau) {
    a + b1 * t
  } else {
    a + b1*tau + b2 * (t - tau)
  }
}
pmax <- breakpoint(tau, tau, a, b1, b2)
tibble(t = 0:365) %>% 
  rowwise() %>% 
  mutate(p = breakpoint(t, tau, a, b1, b2)) %>% 
  ggplot(aes(t, p)) +
  geom_line() +
  theme_bw() +
  xlab("Time [day]") + ylab("Photosynthetic capacity") +
  geom_vline(xintercept = tau, linetype = "dashed") +
  geom_hline(yintercept = pmax, linetype = "dashed")
```

```{r}
expand_grid(b1 = seq(0.1, 0.2, 0.05),
            b2 = c(-0.1, -0.05, -0.025),
            tau = c(100, 200, 300),
            time = 0:365) %>% 
  rowwise() %>% 
  mutate(p = breakpoint(time, tau, a = 0, b1, b2)) %>% 
  ggplot(aes(time, p, col = tau, group = tau)) +
  geom_line() +
  theme_bw() +
  xlab("Time [day]") + ylab("Photosynthetic capacity") +
  facet_grid(b1 ~ b2, labeller = label_both)
```

## Data

I will use the same data as before.

```{r}
ggplot(data, aes(age, p)) +
  geom_point() +
  theme_bw()
```

## Fit

```{r}
model <- cmdstan_model("breakpoint.stan")
model$print()
```

```{r}
#| output: false
mdata <- list(
  N = nrow(data),
  t = data$age,
  p = data$p
)
fit <- model$sample(mdata, parallel_chains = 4)
```

```{r}
fit$draws(c("lp__", "a", "b1", "b2", "tau")) %>% mcmc_trace() + theme_bw()
```

```{r}
#| message: false
#| warnign: false
fit$draws(c("tmax", "pmax")) %>% mcmc_dens() + theme_bw()
```

```{r}
data %>% 
  mutate(pred = fit$summary("mu", median)$median) %>% 
  ggplot(aes(pred, p)) +
  geom_abline() +
  geom_point() +
  theme_bw()
```

```{r}
data %>% 
  mutate(pred = fit$summary("mu", median)$median) %>% 
  ggplot(aes(age, p)) +
  geom_line(aes(y = pred)) +
  geom_point() +
  theme_bw()
```
