---
title: "Bertille 2"
author: "S. Schmitt"
date: last-modified
format: 
  html:
    embed-resources: true
    toc: true
bibliography: references.bib
execute:
  echo: false
---

# Links

-   Bruno's article: <https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/j.1365-2745.2011.01883.x>
-   Ecology letters' article: <https://onlinelibrary.wiley.com/doi/full/10.1111/ele.14318>
-   `cmdstanr` installation: <https://mc-stan.org/cmdstanr/articles/cmdstanr.html>
-   `stan` programming: <https://mc-stan.org/>

# Setup

You only need the `tidyverse` package for table manipulation and figures preparation, `cmdstanr`, and `bayesplot` for Bayesian statistics.

```{r libs}
#| message: false
#| warning: false
library(tidyverse)
library(dplyr)
library(cmdstanr)
library(bayesplot)
```

# Data

I'm reloading the data myself as I'm lacking files to exploit your code.

```{r load}
#| message: false
#| warning: false
data <- readxl::read_xlsx("data/Gx inventaire ALL propre Avril 2025 - IM.xlsx") %>% 
  select(idArbre, "Genre espèce homogènes DB",
         starts_with("C"), starts_with("Vivant")) %>%
  select(-codeArbre, -codeDB, -"COMMENT recru", -"COMMENT mortalité") %>% 
  rename(id = idArbre) %>% 
  separate("Genre espèce homogènes DB", c("genus", "species")) %>% 
  gather(year, value, -id, -genus, -species) %>% 
  mutate(variable = ifelse(str_sub(year, 1, 1) == "C", "circ", "alive")) %>% 
  mutate(year = gsub("C", "", year)) %>% 
  mutate(year = gsub("Vivant", "", year)) %>% 
  mutate(year = as.numeric(year)) %>% 
  pivot_wider(names_from = variable, values_from = value) %>% 
  mutate(circ = as.numeric(circ), alive = as.logical(alive)) %>% 
  filter(alive) %>% 
  mutate(dbh = circ/pi) %>% 
  select(genus, species, id, year, dbh) %>% 
  na.omit() %>% 
  group_by(species) %>% 
  mutate(dbh_max = quantile(dbh, .95))
data %>% 
  group_by(genus, species) %>% 
  summarise(n = length(unique(id)),
            y_min = min(year),
            y_max = max(year),
            dbh_min = min(dbh),
            dbh_max = max(dbh)) %>% 
  arrange(desc(n)) %>% 
  head() %>% 
  knitr::kable()
```

I will focus first on *Eperua falcata* trees.

```{r}
data_sub <- filter(data, genus == "Eperua", species == "falcata")
ggplot(data_sub, aes(year, dbh, group =  id)) +
  geom_line() +
  theme_bw()
```

Then I'm looking only at recruited tress for a first test.

```{r}
inds <- filter(data_sub, dbh < 11) %>% 
  group_by(genus, species, id) %>% 
  summarise(y0 = min(year))
data_sub_rec <- data_sub %>% 
  filter(id %in% inds$id) %>% 
  left_join(inds) %>% 
  mutate(y = year - y0 + 1)
ggplot(data_sub_rec, aes(y, dbh, col = as.character(id))) +
  geom_line() +
  theme_bw()
```

# Single tree

Focusing on a signle recruited *Eperua flacata:*

```{r}
data_ind <- data_sub_rec %>% 
  filter(id == 224946)
ggplot(data_ind, aes(y, dbh, col = as.character(id))) +
  geom_line() +
  theme_bw()
```

$$
DBH(t) \sim \mathcal N(DBH_0+\sum_0^tGmax\times exp(-\frac12[\frac{log(DBH/Dopt)}{K}]^2),\sigma)
$$

```{r }
model <- cmdstan_model("models/sylvain_tree.stan")
model$print()
```

```{r }
#| eval: false
mdata <- list(
  N = nrow(data_ind[-1,]),
  Y = max(data_ind$y),
  dbh0 = data_ind$dbh[1],
  dmax = unique(data_ind$dbh_max),
  dbh = data_ind$dbh[-1],
  year = data_ind$y[-1]
)
fit <- model$sample(data = mdata, parallel_chains = 4, output_dir = "chains/tree")
```

```{r}
fit <- as_cmdstan_fit(list.files("chains/tree/", full.names = TRUE))
```

```{r }
fit$draws(c("lp__", "gmax", "dopt", "k")) %>% 
  mcmc_trace() +
  theme_bw()
```

We summarise the parameter posteriors.

```{r sylvain_tab}
fit$summary(c("gmax", "dopt", "k", "sigma")) %>% 
  knitr::kable()
```

We draw the posterior distributions of the parameters.

```{r }
fit$draws(c("gmax", "dopt", "k", "sigma")) %>% 
  mcmc_dens() +
  theme_bw()
```

```{r }
bind_cols(data_ind[-1,], fit$summary('mu')) %>% 
  ggplot(aes(median, dbh)) +
  geom_abline() +
  geom_segment(aes(x= q5, xend = q95)) +
  geom_point() +
  theme_bw() +
  xlab("Predicted [mm/yr]") + ylab("Observed [mm/yr]")
```

```{r }
bind_cols(data_ind[-1,], fit$summary('mu')) %>% 
  ggplot(aes(y, dbh)) +
  geom_line(aes(y = median)) +
  geom_point() + 
  theme_bw() +
  xlab("Time since recruitment [yr]") + ylab("DBH [cm]")
```

# Single species - recruits

```{r}
data_sub_rec <- data_sub_rec %>% 
  group_by(id) %>% 
  filter(n() > 1) %>% 
  ungroup() %>% 
  mutate(indnum = as.numeric(as.factor(id)))
data_sub_rec %>% 
  ggplot(aes(y, dbh, group = id)) +
  geom_line() +
  theme_bw()
```

For each individual $i$:

$$
\begin{align}
DBH(t,i) \sim \mathcal N(DBH_{0,i}+\sum_0^tGmax_i\times exp(-\frac12[\frac{log(DBH/Dopt_i)}{K_i}]^2),\sigma) \\
Gmax_i \sim \mathcal N(Gmax, \sigma_G) \\
Dopt_i \sim \mathcal N(Dopt, \sigma_D) \\
K_i \sim \mathcal N(K, \sigma_K)
\end{align}
$$

```{r}
model <- cmdstan_model("models/sylvain_species.stan")
model$print()
```

```{r}
subdata <- filter(data_sub_rec, y != 1)
mdata <- list(
  N = nrow(subdata),
  Y = max(subdata$y),
  I = max(subdata$indnum),
  dbh0 = filter(data_sub_rec, y == 1)$dbh,
  dmax = unique(subdata$dbh_max),
  dbh = subdata$dbh,
  year = subdata$y,
  ind = subdata$indnum
)
```

```{r }
#| eval: false
fit <- model$sample(data = mdata, parallel_chains = 4, output_dir = "chains/recruit/")
```

```{r}
fit <- as_cmdstan_fit(list.files("chains/recruit/", full.names = TRUE))
```

> Tree depth should be increased (time consuming).

```{r}
fit$draws(c("lp__", "gmax", "d", "k")) %>% 
  mcmc_trace() +
  theme_bw()
```

```{r}
fit$draws(c("gmax", "d", "k", "sigma")) %>% 
  mcmc_dens() +
  theme_bw()
```

```{r}
fit$draws(c("sigma", "sigma_g", "sigma_d", "sigma_k")) %>% 
  mcmc_dens() +
  theme_bw()
```

```{r}
fit$draws(c("gmax_i")) %>% 
  mcmc_intervals() +
  theme_bw()
```

```{r}
fit$draws(c("d_i")) %>% 
  mcmc_intervals() +
  theme_bw()
```

```{r}
bind_cols(subdata, fit$summary('mu')) %>% 
  ggplot(aes(median, dbh)) +
  geom_abline() +
  geom_segment(aes(x= q5, xend = q95)) +
  geom_point() +
  theme_bw() +
  xlab("Predicted [mm/yr]") + ylab("Observed [mm/yr]")
```

```{r}
bind_cols(subdata, fit$summary('mu')) %>% 
  ggplot(aes(y, dbh, group = id)) +
  geom_line(aes(y = median)) +
  geom_point() + 
  theme_bw() +
  xlab("Time since recruitment [yr]") + ylab("DBH [cm]")
```

# Single species - all

> ⚠️ Long!

```{r}
data_sub <- data_sub %>% 
  group_by(id) %>% 
  filter(n() > 1) %>% 
  mutate(y0 = min(year)) %>% 
  ungroup() %>% 
  mutate(y = year - y0 + 1) %>% 
  mutate(indnum = as.numeric(as.factor(id)))
data_sub %>% 
  ggplot(aes(y, dbh, group = id)) +
  geom_line() +
  theme_bw()
```

For each individual $i$:

$$
\begin{align}
DBH(t,i) \sim \mathcal N(DBH_{0,i}+\sum_0^tGmax_i\times exp(-\frac12[\frac{log(DBH/Dopt_i)}{K_i}]^2),\sigma) \\
Gmax_i \sim \mathcal N(Gmax, \sigma_G) \\
Dopt_i \sim \mathcal N(Dopt, \sigma_D) \\
K_i \sim \mathcal N(K, \sigma_K)
\end{align}
$$

```{r}
model <- cmdstan_model("models/sylvain_species.stan")
model$print()
```

```{r}
subdata <- filter(data_sub, y != 1)
mdata <- list(
  N = nrow(subdata),
  Y = max(subdata$y),
  I = max(subdata$indnum),
  dbh0 = filter(data_sub, y == 1)$dbh,
  dmax = unique(subdata$dbh_max),
  dbh = subdata$dbh,
  year = subdata$y,
  ind = subdata$indnum
)
```

```{r }
#| eval: false
fit <- model$sample(data = mdata, parallel_chains = 4, output_dir = "chains/species/")
```

```{r}
fit <- as_cmdstan_fit(list.files("chains/species/", full.names = TRUE))
```

> Tree depth should be increased (time consuming).

```{r}
fit$draws(c("lp__", "gmax", "d", "k")) %>% 
  mcmc_trace() +
  theme_bw()
```

```{r}
fit$draws(c("gmax", "d", "k", "sigma")) %>% 
  mcmc_dens() +
  theme_bw()
```

```{r}
fit$draws(c("sigma", "sigma_g", "sigma_d", "sigma_k")) %>% 
  mcmc_dens() +
  theme_bw()
```

```{r}
fit$draws(c("gmax_i")) %>% 
  mcmc_intervals() +
  theme_bw()
```

> I think the big gmax_i are possible errors.

```{r}
fit$draws(c("d_i")) %>% 
  mcmc_intervals() +
  theme_bw()
```

```{r}
bind_cols(subdata, fit$summary('mu')) %>% 
  ggplot(aes(median, dbh)) +
  geom_abline() +
  geom_segment(aes(x= q5, xend = q95)) +
  geom_point() +
  theme_bw() +
  xlab("Predicted [mm/yr]") + ylab("Observed [mm/yr]")
```

```{r}
bind_cols(subdata, fit$summary('mu')) %>% 
  ggplot(aes(y, dbh, group = id)) +
  geom_line(aes(y = median)) +
  geom_point() + 
  theme_bw() +
  xlab("Time since recruitment [yr]") + ylab("DBH [cm]")
```

# Multi-species - recruits

```{r}
data_sub <- data %>% 
  ungroup() %>% 
  filter((genus == "Eperua" & species == "falcata") |
           (genus == "Licania" & species == "alba"))
inds <- filter(data_sub, dbh < 11) %>% 
  group_by(genus, species, id) %>% 
  summarise(y0 = min(year))
data_sub_rec <- data_sub %>% 
  filter(id %in% inds$id) %>% 
  left_join(inds) %>% 
  mutate(y = year - y0 + 1) %>% 
  group_by(id) %>% 
  filter(n() > 1) %>% 
  ungroup() %>% 
  mutate(indnum = as.numeric(as.factor(id))) %>% 
  mutate(speciesnum = as.numeric(as.factor(paste(genus, species))))
ggplot(data_sub_rec, aes(y, dbh, group = id, col = species)) +
  geom_line() +
  theme_bw()
```

For each individual $i$ of species $s$:

$$
\begin{align}
DBH(t,i) \sim \mathcal N(DBH_{0,i}+\sum_0^tGmax_i\times exp(-\frac12[\frac{log(DBH/Dopt_i)}{K_i}]^2),\sigma) \\
Gmax_i \sim \mathcal N(Gmax_s, \sigma_G) \\
Dopt_i \sim \mathcal N(Dopt_s, \sigma_D) \\
K_i \sim \mathcal N(K_s, \sigma_K)
\end{align}
$$

> ⚠️ there is a strong hypothesis behind that variation of individual growth from one species to another is similar (community value of sigmaG).

```{r}
model <- cmdstan_model("models/sylvain_multispecies.stan")
model$print()
```

```{r}
subdata <- filter(data_sub_rec, y != 1)
mdata <- list(
  N = nrow(subdata),
  Y = max(subdata$y),
  I = max(subdata$indnum),
  S = max(subdata$speciesnum),
  dbh0 = filter(data_sub_rec, y == 1)$dbh,
  dmax = (select(subdata, speciesnum, dbh_max) %>% unique() %>% arrange(speciesnum))$dbh_max,
  dbh = subdata$dbh,
  year = subdata$y,
  ind = subdata$indnum,
  indsp = (select(subdata, indnum, speciesnum) %>% unique() %>% arrange(indnum))$speciesnum
)
```

```{r }
#| eval: false
fit <- model$sample(data = mdata, parallel_chains = 4, output_dir = "chains/two_species/")
```

```{r}
fit <- as_cmdstan_fit(list.files("chains/two_species/", full.names = TRUE))
```

> Tree depth should be increased (time consuming).

```{r}
fit$draws(c("lp__", "gmax", "d", "k")) %>% 
  mcmc_trace() +
  theme_bw()
```

```{r}
fit$draws(c("gmax", "d", "k", "sigma")) %>% 
  mcmc_dens() +
  theme_bw()
```

> *Licania* growth is small than *Eperua*.

```{r}
fit$draws(c("sigma", "sigma_g", "sigma_d", "sigma_k")) %>% 
  mcmc_dens() +
  theme_bw()
```

```{r}
fit$draws(c("gmax_i")) %>% 
  mcmc_intervals() +
  theme_bw()
```

```{r}
fit$draws(c("d_i")) %>% 
  mcmc_intervals() +
  theme_bw()
```

```{r}
bind_cols(subdata, fit$summary('mu')) %>% 
  ggplot(aes(median, dbh, col = paste(genus, species))) +
  geom_abline() +
  geom_segment(aes(x= q5, xend = q95)) +
  geom_point() +
  theme_bw() +
  xlab("Predicted [mm/yr]") + ylab("Observed [mm/yr]")
```

```{r}
bind_cols(subdata, fit$summary('mu')) %>% 
  ggplot(aes(y, dbh, group = id, col = paste(genus, species))) +
  geom_line(aes(y = median)) +
  geom_point() + 
  theme_bw() +
  xlab("Time since recruitment [yr]") + ylab("DBH [cm]")
```
